Title Text

### 2011-06-09 talk/iwtc11/paper.tex

\begin{figure}
\begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
# arguments to the trace: $p_{0}$, $p_{1}$
# inside f: y = y.add(step)
guard_class($p_{1}$, BoxedInteger)
    # inside BoxedInteger.add
    $i_{2}$ = get($p_{1}$, intval)
    guard_class($p_{0}$, BoxedInteger)
        # inside BoxedInteger.add__int
        $i_{3}$ = get($p_{0}$, intval)
        $i_{4}$ = int_add($i_{2}$, $i_{3}$)
        $p_{5}$ = new(BoxedInteger)
            # inside BoxedInteger.__init__
            set($p_{5}$, intval, $i_{4}$)
jump($p_{0}$, $p_{5}$)
\end{lstlisting}
\caption{An Unoptimized Trace of the Example Interpreter}
\label{fig:unopt-trace}
\end{figure}

@@

@@

>>>>

\begin{figure}
\begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
$l_0$($p_{0}$, $p_{1}$):
# inside f: y = y.add(step)
guard_class($p_{1}$, BoxedInteger)
    # inside BoxedInteger.add
    $i_{2}$ = get($p_{1}$, intval)
    guard_class($p_{0}$, BoxedInteger)
        # inside BoxedInteger.add__int
        $i_{3}$ = get($p_{0}$, intval)
        $i_{4}$ = int_add($i_{2}$, $i_{3}$)
        $p_{5}$ = new(BoxedInteger)
            # inside BoxedInteger.__init__
            set($p_{5}$, intval, $i_{4}$)
jump($l_0$, $p_{0}$, $p_{5}$)
\end{lstlisting}
\caption{An Unoptimized Trace of the Example Interpreter}
\label{fig:unopt-trace}
\end{figure}

@@

The label of the loop is $l_0$ and is used by the jump instruction to
identify it's jump target.

@@

\section{Optimizations}
Before the trace is passed to a backend compiling it into machine code
it needs to be optimized to achieve better performance.
The focus of this paper
is loop invariant code motion. The goal of that is to move as many
operations as possible out of the loop making them executed only once
and not every iteration. This we propose to achieve by loop peeling. It
leaves the loop body intact, but prefixes it with one iteration of the
loop. This operation by itself will not achieve anything. But if it is
combined with other optimizations it can increase the effectiveness of
those optimizations. For many optimization of interest some care has
to be taken when they are combined with loop peeling. This is
described below by first explaining the loop peeling optimization
followed by a set of other optimizations and how they interact with
loop peeling.

\subsection{Loop peeling}
Loop peeling is achieved by inlining the trace at the end of
itself. The input arguments of the second iteration are replaced with
the jump arguments of the first iterations and then the arguments of all
the operations are updated to operate on the new input arguments. To
keep the single-assignment form new variables has to be introduced as
the results of all the operations. The first iteration of the loop
will end with a jump to the second iteration of the loop while the
second iteration will end with a jump to itself. This way the first
copy of the trace only be executed once while the second copy will be
used for every other iteration. The rationality here is that the
optimizations below typically will be able to optimize the second copy
more efficiently than the first. The trace from Figure~\ref{fig:unopt-trace} would
after this operation become the trace in Figure~\ref{fig:peeled-trace}.

\begin{figure}
\begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
$l_0$($p_{0}$, $p_{1}$):
# inside f: y = y.add(step)
guard_class($p_{1}$, BoxedInteger)
    # inside BoxedInteger.add
    $i_{2}$ = get($p_{1}$, intval)
    guard_class($p_{0}$, BoxedInteger)
        # inside BoxedInteger.add__int
        $i_{3}$ = get($p_{0}$, intval)
        $i_{4}$ = int_add($i_{2}$, $i_{3}$)
        $p_{5}$ = new(BoxedInteger)
            # inside BoxedInteger.__init__
            set($p_{5}$, intval, $i_{4}$)
jump($l_1$, $p_{0}$, $p_{5}$)

$l_1$($p_{0}$, $p_{5}$):
# inside f: y = y.add(step)
guard_class($p_{5}$, BoxedInteger)
    # inside BoxedInteger.add
    $i_{6}$ = get($p_{5}$, intval)
    guard_class($p_{0}$, BoxedInteger)
        # inside BoxedInteger.add__int
        $i_{7}$ = get($p_{0}$, intval)
        $i_{8}$ = int_add($i_{6}$, $i_{7}$)
        $p_{9}$ = new(BoxedInteger)
            # inside BoxedInteger.__init__
            set($p_{9}$, intval, $i_{8}$)
jump($l_1$, $p_{0}$, $p_{9}$)
\end{lstlisting}
\caption{An Unoptimized Trace of the Example Interpreter}
\label{fig:peeled-trace}
\end{figure}

When applying the following optimizations to this two iteration trace
some care has to taken as to how the jump arguments of both
iterations and the input arguments of the second iteration are
treated. It has to be ensured that the second iteration stays a proper
trace in the sens that the operations within it only operations on
variables that are either among the input arguments of the second iterations
or are produced within the second iterations. To ensure this we need
to introduce a bit of formalism.

The original trace (prior too peeling) consists of three parts. 
A vector of input
variables, $I=\left(I_1, I_2, \cdots, I_{|I|}\right)$, a list of non
jump operations and a single
jump operation. The jump operation contains a vector of jump variables,
$J=\left(J_1, J_2, \cdots, J_{|J|}\right)$, that are passed as the input variables of the target loop. After
loop peeling there will be a second copy of this trace with input
variables equal to the jump arguments of the first copy, $J$, and jump
arguments $K$. Looking back at our example we have
\begin{equation}
  %\left\{
    \begin{array}{lcl}
      I &=& \left( p_0, p_1 \right) \\
      J &=& \left( p_0, p_5 \right) \\
      K &=& \left( p_0, p_9 \right) \\
    \end{array}
  %\right.
  .
\end{equation}
To construct the second iteration from the first we also need a
function, $m$, mapping the variables of the first iteration onto the
variables of the second. This function is constructed during the
inlining. It is initialized by mapping the input arguments, $I$, to
the jump arguments $J$,
\begin{equation}
  m\left(I_i\right) = J_i \ \text{for}\ i = 1, 2, \cdots |I| .
\end{equation}
In the example that means (XXX which notation do we prefer?)
\begin{equation}
  m(v) = 
  \left\{
    \begin{array}{lcl}
      p_0 &\text{if}& v=p_0 \\
      p_5 &\text{if}& v=p_1 \\
    \end{array}
  \right.
  .
\end{equation}
\begin{equation}
  %\left\{
    \begin{array}{lcl}
      m\left(p_0\right) &=& p_0 \\
      m\left(p_1\right) &=& p_5
    \end{array}
  %\right.
  .
\end{equation}
Each operation in the trace is inlined in the order they are
executed. To inline an operation with argument vector 
$A=\left(A_1, A_2, \cdots, A_{|A|}\right)$ producing the variable $v$
a new variable, $\hat v$ is introduced. The inlined operation will
produce $\hat v$ from the input arguments 
\begin{equation}
  \left(m\left(A_1\right), m\left(A_2\right), 
    \cdots, m\left(A_{|A|}\right)\right) . 
\end{equation}
Before the
next operation is inlined, $m$ is extend by making $m\left(v\right) = \hat
v$. After all the operations in the example have been inlined we have
\begin{equation}
  %\left\{
    \begin{array}{lcl}
      m\left(p_0\right) &=& p_0 \\
      m\left(p_1\right) &=& p_5 \\
      m\left(i_2\right) &=& i_6 \\
      m\left(i_3\right) &=& i_7 \\
      m\left(i_4\right) &=& i_8 \\
      m\left(p_5\right) &=& p_9 \\
    \end{array}
  %\right.
  .
\end{equation}

\subsection{Redundant guard removal}
No special concerns needs to be taken when implementing redundant
guard removal together with loop peeling. However the the guards from
the first iteration might make the guards of the second iterations
redundant and thus removed. So the net effect of combining redundant
guard removal with loop peeling is that guards are moved out of the
loop. The second iteraton of the example reduces to

\begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
$l_1$($p_{0}$, $p_{5}$):
# inside f: y = y.add(step)
    # inside BoxedInteger.add
    $i_{6}$ = get($p_{5}$, intval)
        # inside BoxedInteger.add__int
        $i_{7}$ = get($p_{0}$, intval)
        $i_{8}$ = int_add($i_{6}$, $i_{7}$)
        $p_{9}$ = new(BoxedInteger)
            # inside BoxedInteger.__init__
            set($p_{9}$, intval, $i_{8}$)
jump($l_1$, $p_{0}$, $p_{9}$)
\end{lstlisting}

\subsection{Heap caching}

To implement heap caching variables has to be passed from the first
iteration to the second by XXX
\begin{equation}
  \hat J = \left(J_1, J_2, \cdots, J_{|J|}, H_1, H_2, \cdots, H_{|H}\right)
\end{equation}
\begin{equation}
  \hat K = \left(K_1, K_2, \cdots, K_{|J|}, m(H_1), m(H_2), \cdots, m(H_{|H})\right)
  .
\end{equation}
In the optimized trace $I$ is replaced by $\hat I$ and $K$ by $\hat K$.

\begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
$l_0$($p_{0}$, $p_{1}$):
# inside f: y = y.add(step)
guard_class($p_{1}$, BoxedInteger)
    # inside BoxedInteger.add
    $i_{2}$ = get($p_{1}$, intval)
    guard_class($p_{0}$, BoxedInteger)
        # inside BoxedInteger.add__int
        $i_{3}$ = get($p_{0}$, intval)
        $i_{4}$ = int_add($i_{2}$, $i_{3}$)
        $p_{5}$ = new(BoxedInteger)
            # inside BoxedInteger.__init__
            set($p_{5}$, intval, $i_{4}$)
jump($l_1$, $p_{0}$, $p_{5}$, $i_3$, $i_4$)

$l_1$($p_{0}$, $p_{5}$, $i_3$, $i_4$):
# inside f: y = y.add(step)
    # inside BoxedInteger.add
        # inside BoxedInteger.add__int
        $i_{8}$ = int_add($i_{4}$, $i_{3}$)
        $p_{9}$ = new(BoxedInteger)
            # inside BoxedInteger.__init__
            set($p_{9}$, intval, $i_{8}$)
jump($l_1$, $p_{0}$, $p_{9}$, $i_3$, $i_8$)
\end{lstlisting}

\subsection{Virtualization}
Using escape analysis we can XXX

Let $\tilde J$ be all variables in $J$ not representing virtuals (in the
same order). Extend it with all non virtual fields, $H_i$, of the
removed virtuals,
\begin{equation}
  \hat J = \left(\tilde J_1, \tilde J_2, \cdots, \tilde J_{|\tilde J|}, 
                 H_1, H_2, \cdots, H_{|H}\right)
\end{equation}
and let
\begin{equation}
  \hat K = \left(m\left(\hat J_1\right), m\left(\hat J_1\right), 
                 \cdots, m\left(\hat J_{|\hat J|}\right)\right)
  .
\end{equation}

\begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
$l_0$($p_{0}$, $p_{1}$):
# inside f: y = y.add(step)
guard_class($p_{1}$, BoxedInteger)
    # inside BoxedInteger.add
    $i_{2}$ = get($p_{1}$, intval)
    guard_class($p_{0}$, BoxedInteger)
        # inside BoxedInteger.add__int
        $i_{3}$ = get($p_{0}$, intval)
        $i_{4}$ = int_add($i_{2}$, $i_{3}$)
jump($l_1$, $p_{0}$, $i_3$, $i_4$)

$l_1$($p_{0}$, $p_{5}$, $i_3$, $i_4$):
# inside f: y = y.add(step)
    # inside BoxedInteger.add
        # inside BoxedInteger.add__int
        $i_{8}$ = int_add($i_{4}$, $i_{3}$)
jump($l_1$, $p_{0}$, $i_3$, $i_8$)
\end{lstlisting}

And we're down to a single integer addition!

\section{Benchmarks}
